---
title: "Cultures microbial plate counts"
author: "L. Bettera & M. Dreier"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---


```{r mb packages, include=FALSE}
source('genfuncs.R')
Rdeps = c('openxlsx', 'plyr', 'gridExtra', 'tidyverse', 'outliers', 
          "reshape2", "stats", "factoextra", "FactoMineR", "agricolae", 
          'FSA', 'ggplot2', 'patchwork')
checkpackages(Rdeps)
```


```{r  mb data import}
# Uploading the files and creating subsets
fp <- file.path(getwd(), "..", "data", "cultures", "microbes", "viable_counts.xlsx")
df <- read.xlsx(fp, sheet = 1) # import in R the dataset (excel file) as dataframe

# let's transform the first variable in factor (useful for data exploration by groups)
df <- df %>% mutate_at(vars(sample, replicate),
                       as.factor)
# let's remove the repl. 1 because considered outlier from NGS analysis
df <- subset(df, replicate != "1")
df <- df[-2] # let's remove the "replicate" column
```


We analyzed 12 samples:

1.  RM = raw milk

2.  RM.H = raw milk heat treated (54째C for 54min)

3.  eRM = enriched raw milk (spontaneous fermentation at 10째C for 21d)

4.  eRM.H = enriched heated raw milk

5.  eRM.HS = enriched heated, salted raw milk (5% v/w NaCl)

6.  NWC = natural whey culture (thermophilic)

7.  eRWC.y = enriched raw milk whey culture, young. Mix NWC-eRM (1:10), incubated at 38째C for 6h

8.  eRWC.o = old, incubated at 38째C for 22h

9.  eRWC.H.y = mix NWC-eRM.H (1:10), young

10. eRWC.H.o = mix NWC-eRM.H (1:10), old

11. eRWC.HS.y = mix NWC-eRM.HS (1:10), young

12. eRWC.HS.o = mix NWC-eRM.HS (1:10), old

Here is a summary statistic of the results divided for each sample:

```{r summary statistics}
# a function derived from "R in action" book
mystats <- function(x, na.omit=TRUE){
  if (na.omit)
    x <- x[!is.na(x)]
  n <- round(length(x), digits = 1)
  mx <- round(max(x), digits = 1)
  mn <- round(min(x), digits = 1)
  m <- round(mean(x), digits = 1)
  s <- round(sd(x), digits = 1)
  return(c(n = n, mean = m, stdev = s, max = mx, min = mn))
}

dstats <- function(x)sapply(x, mystats)
dfstats <- by(df[2:9],
              list(sample = df$sample), dstats)
dfstats

```

## Data analysis: RM and eRM

### Boxplot

This boxplot shows the spontaneous fermentation effect of RM on the concentration of different microbial groups.

```{r boxplot1, dpi=150}
# for boxplots, the dataset in melted form is preferred
df2 <- melt(df, id = "sample", value.name = "CFU_mL")

bp1 <-  ggplot(filter(df2, sample %in% c("RM", "eRM")),
                   aes(x = variable,
                       y = CFU_mL,
                       fill = factor(sample, levels = c("RM", "eRM")))) +
        geom_boxplot() +
        scale_y_continuous(breaks = seq(0, 10, by = 1), limits = c(0, 10)) +
        scale_fill_manual(name  = "Sample",
                          breaks = c("RM", "eRM"),
                          labels = c("RM", "eRM"),
                          values = c("#88CCEE", "#CC6677")) +
        theme(axis.title.x = element_blank()) +
        theme(axis.title.y = element_text(size = 12)) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12)) +
        theme(axis.text.y = element_text(size = 14)) +
        theme(panel.background = element_rect(fill = "white")) +
        theme(panel.border = element_blank(),
              panel.grid.major.y = element_line(color = "#C8C8C8"),
              panel.grid.major.x = element_blank(),
              panel.grid.minor = element_blank(),
              axis.line = element_line(colour = "black")) +
        theme(legend.text = element_text(size = 12),
              legend.title = element_text(size = 12)) +
        ylab("log CFU/mL")
bp1
```

## Statistics

Statistical analysis was performed in order to define whether the evaluated differences were significant. If the variances of the two groups resulted homogeneous to the Fisher's test, a Two Sample t-test was applied; alternatively a Welch t-test was used.

```{r testing_hypotheses1}
# variance homogeneity (for interpretation http://www.sthda.com/english/wiki/f-test-compare-two-variances-in-r)
# If p-value>0.05 variances are homogeneous

sample_names <- c("RM", "eRM")

subdf2 <- filter(df2, sample %in% sample_names)
m_vars = unique(subdf2$variable)

for (x in m_vars) {
  test_two_grps(x, subdf2, sample_names)
}
```

## Data analysis: RM.H, eRM.H and eRM.HS 

### Statistics

Statistical analysis was performed in order to define whether the evaluated differences were significant. If the variances of the two groups resulted homogeneous to the Fisher's test, a Two Sample t-test was applied; alternatively a Welch t-test was used.

```{r testing_hypotheses2}
sample_names <- c("RM.H", "eRM.H")

subdf3 <- filter(df2, sample %in% sample_names)
m_vars = unique(subdf3$variable)

for (x in m_vars) {
  test_two_grps(x, subdf3, sample_names)
}
```

```{r testing_hypotheses3}
sample_names <- c("RM.H", "eRM.HS")

subdf4 <- filter(df2, sample %in% sample_names)
m_vars = unique(subdf4$variable)

for (x in m_vars) {
  test_two_grps(x, subdf4, sample_names)
}
```

### Boxplot

Here is shown the microbial enrichment of RM.H, also with addition of salt.

```{r boxplot2, dpi=150}
bp2 <-  ggplot(filter(df2, sample %in% c("RM.H", "eRM.H", "eRM.HS")),
               aes(x = variable,
               y = CFU_mL,
               fill = factor(sample, levels = c("RM.H", "eRM.H", "eRM.HS")))) +
        geom_boxplot() +
        scale_y_continuous(breaks = seq(0, 10, by = 1), limits = c(0, 10)) +
        scale_fill_manual(name  = "Sample",
                          breaks = c("RM.H", "eRM.H", "eRM.HS"),
                          labels = c("RM.H", "eRM.H", "eRM.HS"),
                          values = c("#DDCC77", "#117733", "#332288")) +
        theme(axis.title.x = element_blank()) +
        theme(axis.title.y = element_text(size = 12)) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12)) +
        theme(axis.text.y = element_text(size = 14)) +
        theme(panel.background = element_rect(fill = "white")) +
        theme(panel.border = element_blank(),
              panel.grid.major.y = element_line(color = "#C8C8C8"),
              panel.grid.major.x = element_blank(),
              panel.grid.minor = element_blank(),
              axis.line = element_line(colour = "black")) +
        theme(legend.text = element_text(size = 12),
              legend.title = element_text(size = 12)) +
        ylab("log CFU/mL")
bp2
```
```{r overview boxplot, dpi=150}
selected_samples = c("RM", "RM.H", "eRM", "eRM.H", "eRM.HS")

over_bp <-  ggplot(filter(df2, sample %in% selected_samples),
                   aes(x = variable,
                       y = CFU_mL,
                       fill = factor(sample, levels = selected_samples))) +
            geom_boxplot() +
            scale_y_continuous(breaks = seq(0, 10, by = 1), limits = c(0, 10)) +
            scale_fill_manual(name  = "Sample",
                              breaks = selected_samples,
                              labels = selected_samples,
                              values = c("#88CCEE", "#DDCC77", "#CC6677", "#117733", "#332288")) +
            theme(axis.title.x = element_blank()) +
            theme(axis.title.y = element_text(size = 12)) +
            theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12)) +
            theme(axis.text.y = element_text(size = 14)) +
            theme(panel.background = element_rect(fill = "white")) +
            theme(panel.border = element_blank(),
                  panel.grid.major.y = element_line(color = "#C8C8C8"),
                  panel.grid.major.x = element_blank(),
                  panel.grid.minor = element_blank(),
                  axis.line = element_line(colour = "black")) +
            theme(legend.text = element_text(size = 12),
                  legend.title = element_text(size = 12)) +
            theme(legend.position = "right",
                  legend.justification = "left",
                  legend.direction = "vertical") + 
            ylab("log CFU/mL")
over_bp
```


### Statistics

Since here there are more then 2 groups, a one-way ANOVA model was needed for the statistical analysis. The variance homogeneity was previously checked using the Bartlett's test. In case the assumptions of one-way ANOVA test are not met, a Kruskal-Wallis test is applied. A Multiple Mean Comparison test was eventually performed afterwards.

```{r testing_hypotheses4}
subdf5 <- filter(df2, sample %in% c("RM.H", "eRM.H", "eRM.HS"))
m_vars = unique(subdf5$variable)

for (x in m_vars) {
  test_more_grps(x, subdf5)
}
```


## Data analysis: NWC and final cultures 

### Boxplot

This boxplot shows the microbial composition of the final cultures, obtained by incubating the mixture of NWC and the enriched milks.

```{r boxplot3, dpi=150}
bp3 <-  ggplot(filter(df2, sample %in% c("NWC", "eRWC.y", "eRWC.o", "eRWC.H.y",
                                        "eRWC.H.o", "eRWC.HS.y", "eRWC.HS.o")),
                   aes(x = variable, 
                       y = CFU_mL,
                       fill = factor(sample, levels = c("NWC", "eRWC.y", "eRWC.o", "eRWC.H.y",
                                                        "eRWC.H.o", "eRWC.HS.y", "eRWC.HS.o")))) +
        geom_boxplot(position = position_dodge(0.7)) +
        scale_y_continuous(breaks = seq(0, 10, by = 1), limits = c(0, 10)) +
        scale_fill_manual(name  = "Sample",
                          breaks = c("NWC", "eRWC.y", "eRWC.o", "eRWC.H.y",
                                       "eRWC.H.o", "eRWC.HS.y", "eRWC.HS.o"),
                          labels = c("NWC", "eRWC.y", "eRWC.o", "eRWC.H.y",
                                       "eRWC.H.o", "eRWC.HS.y", "eRWC.HS.o"),
                          values = c("#88CCEE", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499", "#44AA99")) +
        theme(axis.title.x = element_blank()) +
        theme(axis.title.y = element_text(size = 12)) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12)) +
        theme(axis.text.y = element_text(size = 14)) +
        theme(panel.background = element_rect(fill = "white")) +
        theme(panel.border = element_blank(),
              panel.grid.major.y = element_line(color = "#C8C8C8"),
              panel.grid.major.x = element_blank(),
              panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) +
        theme(legend.text = element_text(size = 12),
              legend.title = element_text(size = 12)) +
        theme(legend.position = "right",
              legend.justification = "left",
              legend.direction = "vertical") + 
        ylab("log CFU/mL")
```


```{r summary figure, fig.width=12,fig.height=10}
summary_fig <- over_bp + bp3 + plot_layout(nrow= 2) + plot_annotation(tag_levels = "A") &
            theme(plot.tag = element_text(size = 18, face = "bold"))
summary_fig

```


## Testing hypotheses

Since here there are more then 2 groups, a one-way ANOVA model was needed for the statistical analysis. The variance homogeneity was previously checked using the Bartlett's test. In case the assumptions of one-way ANOVA test are not met, a Kruskal-Wallis test is applied. A Multiple Mean Comparison test was eventually performed afterwards.

```{r testing_hypotheses5}
selected_samples <- c("NWC", "eRWC.y", "eRWC.o", "eRWC.H.y", "eRWC.H.o", "eRWC.HS.y", "eRWC.HS.o")

subdf6 <- filter(df2, sample %in% selected_samples)
m_vars = unique(subdf6$variable)

for (x in m_vars) {
  test_more_grps(x, subdf6)
}
```

